---
title: "VABHS Assessment"
author: "Jeffrey Letourneau"
date: "`r Sys.Date()`"
output:
  html_document:
    df_print: paged
---

# Purpose
Build and evaluate a predictive model of one-year survival after diagnosis with Non-small cell lung cancer (NSCLC).

# Load libraries
```{r, results = F}
# tidyverse contains dplyr for manipulating dataframes, ggplot2 for plotting, and other useful packages
library(tidyverse) 
library(RColorBrewer)

# Packages used for random forest model
library(randomForest)
library(caret)
library(ROCR)
library(pROC)
library(MLmetrics)
library(MLeval)

# Set preferred ggplot theme
theme_set(theme_bw() +
            theme(axis.text = element_text(color = "black"),
                  axis.ticks = element_line(color = "black"),
                  plot.title = element_text(hjust = 0.5)))

# Set seed for reproducibility
set.seed(123)

```

# Load and inspect data
```{r}
clinical <- read.csv("data/clinical.csv")
genomics <- read.csv("data/genomics.csv")

# Merge datasets
## Make genomics in same format as clinical (participants are rows, variables are columns) 
genomics <- genomics %>%
  mutate(value = "mutation") %>%
  spread(key = Gene, value = value)
  
## Merge by ID
merge_df <- clinical %>%
  left_join(genomics) %>%
  replace(is.na(.), "WT")

# Inspect for possible errors/duplicates
str(merge_df[,1:16])
n_distinct(merge_df$ID) == nrow(merge_df) # TRUE --> good, no duplicate participants
unique(clinical$Outcome) # "Alive" or "Dead"

## These seem like places typos could have happened
unique(merge_df$`T`) # no issues
unique(merge_df$Stage) # there is a "1B" that should be "IB"
unique(merge_df$Primary.Site) # "Righ Upper Lobe" should be "Right Upper Lobe"
unique(merge_df$Histology) # no issues

## Other changes needed:
### "UNK" and "NULL" --> NA for T, N, M, Tumor.Size
### Make Radiation a categorical variable instead of numeric

# Clean up the merged dataframe typos are corrected and missing data is treated as NA
merge_df <- merge_df %>%
  mutate(Stage = case_when(Stage == "1B" ~ "IB", TRUE ~ Stage),
         Primary.Site = case_when(Primary.Site == "Righ Upper Lobe" ~ "Right Upper Lobe",
                                  TRUE ~ Primary.Site),
         Radiation = as.character(Radiation),
         Radiation = case_when(Radiation == 0 ~ "no", Radiation == 5 ~ "yes"),
         `T` = na_if(`T`, "UNK"),
         N = na_if(N, "NULL"),
         N = as.numeric(N),
         M = na_if(M, "NULL"),
         M = as.numeric(M),
         Tumor.Size = na_if(Tumor.Size, "NULL"))

merge_df

```

# Revising and reducing features
```{r}
# Add new variables
# one_year_survival: categorical variable to represent survival at one year as a binary
# total_meta: total number of metastases (sum of N + M)
# total_tumors: sum of N + M + Num.Primaries

merge_df <- merge_df %>%
  mutate(one_year_survival = case_when(Survival.Months > 12 ~ "Survived",
                                       Survival.Months <= 12 ~ "Deceased"))

merge_df$total_meta <- rowSums(merge_df[,c("N", "M")], na.rm = T)
merge_df$total_tumors <- rowSums(merge_df[,c("Num.Primaries", "N", "M")], na.rm = T)

merge_df <- merge_df %>%
  relocate(one_year_survival:total_tumors, .after = Outcome)


# Reduce features for model
## Remove genes without mutations in at least 10% of participants
cutoff <- 0.1 * nrow(clinical)

genes_to_keep <- read.csv("data/genomics.csv") %>%
  group_by(Gene) %>%
  summarize(count = length(Gene)) %>%
  filter(count > cutoff) %>%
  pull(Gene)

merge_df <- merge_df %>%
  dplyr::select(ID, one_year_survival, total_meta, total_tumors, Age:Num.Mutations, all_of(genes_to_keep))


```

# Handle missing data
There are two main ways we can handle this: (a) remove participants with NAs, or (b) impute NA values. Neither is ideal - (b) essentially sets NAs to the median/mode value for each feature.
```{r}
# How many rows have no NAs
nrow(merge_df[complete.cases(merge_df),]) # only 33, not good!


# Random forest needs everything to be numeric or factor
merge_df_for_rf <- merge_df[,c(3:ncol(merge_df), 2)] %>%
  mutate(one_year_survival = factor(one_year_survival),
         Age = as.numeric(Age),
         Grade = as.numeric(Grade),
         Num.Primaries = as.numeric(Num.Primaries),
         `T` = factor(`T`),
         Radiation = factor(Radiation),
         Stage = factor(Stage),
         Primary.Site = factor(Primary.Site),
         Histology = factor(Histology),
         Tumor.Size = as.numeric(Tumor.Size),
         Num.Mutated.Genes = as.numeric(Num.Mutated.Genes),
         Num.Mutations = as.numeric(Num.Mutations),
         CDKN2A = factor(CDKN2A),
         KRAS_Col1 = factor(KRAS_Col1),
         MSH2 = factor(MSH2),
         STK11 = factor(STK11),
         TP53_Col1 = factor(TP53_Col1),
         TSC2 = factor(TSC2))

merge_df_for_rf <- rfImpute(one_year_survival ~ ., merge_df_for_rf)
  

```


# Build the random forest model
```{r}

# create training/testing split
data <- merge_df_for_rf

# Train the model
fitControl <- trainControl(method = "LOOCV",
                           summaryFunction = prSummary,
                           classProbs = T,
                           savePredictions = TRUE) 
rf <- train(one_year_survival ~ ., data = data,
            method = "rf",
            trControl = fitControl,
            tuneGrid = expand.grid(.mtry=sqrt(ncol(data)-1)), # default val of sqrt(# features)
            verbose = FALSE)

# Store the scaled importance values
importance_df <- varImp(rf)$importance %>% as.matrix()

# Summarize confusion matrix
confusion <- confusionMatrix(data = rf$pred$pred, reference = rf$pred$obs,
                             mode = "prec_recall", positive = "Survived")

# Pull resulting metrics
accuracy <- confusion$overall["Accuracy"]
kappa <- confusion$overall["Kappa"]
pval <- confusion$overall["AccuracyPValue"]

# Plotting ROC curve
plots <- evalm(rf)
roc.data <- data.frame(SENS=plots$roc$data$SENS, FPR=plots$roc$data$FPR,
                           point=1:length(plots$roc$data$SENS))
auc <- plots$stdres$`Group 1`[13,1]

```

# Plot results
```{r}
# ROC curve
roc_plot <- ggplot(roc.data, aes(x=FPR, y=SENS)) +
  geom_line(color="red") +
  labs(x="False Positive Rate", y="True Positive Rate") +
  geom_abline(intercept = 0, slope = 1, color="darkgray") +
  annotate(geom="text", x=0.7, y=0.25, label=paste0("AUC = ", round(auc, 3), "\n",
                                                    "Accuracy = ", round(accuracy, 3), "\n",
                                                    "Kappa = ", round(kappa,3), "\n",
                                                    "Mean p(Acc < NIR) = ", round(pval, 3)))
#ggsave("plots/roc_curve.png", roc_plot, height = 3, width = 4)

roc_plot

# Confusion matrix
confusion_plot <- ggplot(as.data.frame(confusion$table), aes(x = Prediction, y = Reference)) +
  geom_tile(aes(fill = Freq)) +
  geom_text(aes(label = Freq), size = 10) +
  scale_fill_distiller(palette = "RdYlGn", direction = 1) +
  theme(legend.position = "none") +
  labs(x = "Predicted one-year survival", y = "Actual one-year survival")
#ggsave("plots/confusion_matrix.png", confusion_plot, height = 3, width = 4)

confusion_plot

# Plot most important features
top10 <- importance_df %>% 
  as.data.frame() %>%
  arrange(desc(Overall)) %>%
  slice_max(Overall, n = 10) %>%
  mutate(feature = rownames(.)) %>%
  mutate(feature = factor(feature, levels = rev(feature)))

importance_plot <- ggplot(top10, aes(x=Overall, y=feature)) +
  geom_col() +
  labs(x="Importance", y = "Feature")
#ggsave("plots/importance_top10.png", importance_plot, height = 3, width = 4)

importance_plot

```

# Plot individual features
```{r}
# Here, I operate off the original (non-imputed) data

M_plot <- ggplot(merge_df, aes(x = factor(M), y = one_year_survival)) +
  geom_point(position = position_jitter(height = 0.2, width = 0.2), alpha = 0.5) +
  labs(x = "Number of distant metastases (M)", y = "One-year survival")

#ggsave("plots/M.png", M_plot, height = 3, width = 6)

M_plot

hist_plot <- ggplot(merge_df, aes(x = Histology, y = one_year_survival)) +
  geom_point(position = position_jitter(height = 0.2, width = 0.2), alpha = 0.5) +
  labs(x = "Tumor histology", y = "One-year survival")
#ggsave("plots/histology.png", hist_plot, height = 3, width = 6)

hist_plot

```


# Print session info
```{r}
sessionInfo()
```

